Hazelnut: A Bidirectionally Typed \\ Structure Editor Calculus

### 2016-06-22 hazelnut-popl17.tex
added TODOs

\def\OPTIONConf{0}%
%\documentclass[12pt]{article}
\documentclass[preprint,9pt]{sigplanconf}
% The following \documentclass options may be useful:
%
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{ stmaryrd }
\usepackage{mathpartir}
\usepackage{stmaryrd}
\usepackage{wasysym}
\usepackage{extarrows}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\usepackage{mathpazo}
\usepackage{colortab}
\usepackage{url}
\usepackage{todonotes}
\usepackage{listings}
\usepackage{microtype}
\sloppy
\def \TirNameStyle #1{\small\rulename{#1}}
\renewcommand{\MathparLineskip}{\lineskiplimit=.3\baselineskip\lineskip=.3\baselineskip plus .2\baselineskip}

@@

\section{Hazelnut, Formally}
\label{sec:hazel}
Hazelnut is based on the simply-typed lambda calculus extended with a single base type, $\tnum$. Its major constituents, introduced by example in the previous section, are:
\begin{itemize}
\item \textbf{H-types} and \textbf{H-expressions} (Sec. \ref{sec:holes}), which are terms with \emph{holes}. Holes mark subterms that are ``under construction.'' H-types classify H-expressions according to a {bidirectionally typed} static semantics.
\item \textbf{Z-types} and \textbf{Z-expressions} (Sec. \ref{sec:cursors}), which superimpose a single \emph{focus} onto H-types and H-expressions (using Huet's \emph{zipper pattern} \cite{JFP::Huet1997}.)
\item \textbf{Actions} (Sec. \ref{sec:actions}), which move the focus or modify the subterm in focus.

@@

We refer to terms that do not contain subterms of hole form as \emph{complete}. Informally, we will use metavariables $\tau$ and $e$ rather than $\htau$ and $\hexp$ for complete H-types and H-expressions, respectively. Formally, we can derive $\hcomplete{\tau}$ when $\tau$ is a complete H-type, and $\hcomplete{e}$ when $e$ is a complete H-expression. We omit the straightforward definitions of these judgements for concision. The dynamics of Hazelnut, which we need not detail here, is defined only  over complete H-expressions (i.e. we can only ``run'' a complete program, though see Sec. \ref{sec:future}.)

@@

The rules for relative movement within Z-expressions are similar. Movement is type-independent, so we defer to an auxiliary judgement for both the analytic and synthetic judgements:
\begin{subequations}
\begin{equation}
\inferrule{
  \performMove{\zexp}{\aMove{\delta}}{\zexp'}
}{
  \performSyn{\hGamma}{\zexp}{\htau}{\aMove{\delta}}{\zexp'}{\htau}
}
\end{equation}
\begin{equation}
  \inferrule{
  \performMove{\zexp}{\aMove{\delta}}{\zexp'}
}{
  \performAna{\hGamma}{\zexp}{\htau}{\aMove{\delta}}{\zexp'}
}
\end{equation}
\end{subequations}
For concision, we show only the rules for ascription here:
\begin{subequations}
  \begin{equation}
    \label{r:movefirstchild}
  \inferrule{ }{
    \performTyp{
      \zwsel{\hexp : \htau}
    }{
      \aMove{\dChild}
    }{
      \zwsel{\hexp} : \htau
    }
  }
\end{equation}
\begin{equation}
  \label{r:moveparent}
  \inferrule{ }{
    \performTyp{
      \zwsel{\hexp} : \htau
    }{
      \aMove{\dParent}
    }{
      \zwsel{\hexp : \htau}
    }
  }
\end{equation}
\begin{equation}
  \inferrule{ }{
    \performTyp{
      \hexp : \zwsel{\htau}
    }{
      \aMove{\dParent}
    }{
      \zwsel{\hexp : \htau}
    }
  }
\end{equation}
\begin{equation}
  \label{r:movenextsib}
  \inferrule{ }{
    \performTyp{
      \zwsel{\hexp} : \htau
    }{
      \aMove{\dNext}
    }{
      \hexp : \zwsel{\htau}
    }
  }
\end{equation}
\begin{equation}
  \label{r:moveprevsib}
  \inferrule{ }{
    \performTyp{
      \hexp : \zwsel{\htau}
    }{
      \aMove{\dPrev}
    }{
      \zwsel{\hexp} : \htau
    }
  }
\end{equation}
\begin{equation}
\inferrule{
  \performTyp{
    \zexp
  }{
    \aMove{\delta}
  }{
    \zexp'
  }
}{
  \performTyp{
    \zexp : \htau
  }{
    \aMove{\delta}
  }{
    \zexp' : \htau
  }
}
\end{equation}
\begin{equation}
  \inferrule{
    \performTyp{
      \ztau
    }{
      \aMove{\delta}
    }{
      \ztau'
    }
  }{
    \performTyp{
      \hexp : \ztau
    }{
      \aMove{\delta}
    }{
      \hexp : \ztau'
    }
  }
\end{equation}
and this for the example
\begin{equation}\label{r:movefirstchild-lam}
\inferrule{ }{
  \performMove{
    \zwsel{\hlam{x}{\hexp}}
  }{
    \aMove{\dChild}
  }{
    \hlam{x}{\zwsel{\hexp}}
  }
}
\end{equation}

@@

@@

\end{subequations}

@@

\section{Dynamics of Hazelnut}
\label{sec:dynamics}
\input{dynamics}

@@

@@

\section{Discussion \& Conclusion}
\label{sec:future}
This paper presented Hazelnut, a type theoretic structure editor calculus. Our aim is to take a principled approach to its design by formally specifying its semantics, providing strong metatheoretic guarantees, mechanizing its semantics and metatheory in Agda and implementing it using the concepts of  functional reactive programming. As of this submission, we have achieved reasonable confidence in the formal system presented above, and have transitioned our focus toward the mechanization and implementation efforts. By the time of presentation, we anticipate having complete or nearly complete versions of these.

@@

It may also be possible to give a dynamics to incomplete expressions. Prior work on staged evaluation suggests that there may be a connection to modal logic, viewing holes as quantifying over all possible terms that may fill them \cite{DBLP:journals/jacm/DaviesP01}. In developing a dynamic semantics, we will also need to handle terms like $\hhole{\hehole}$ and
$\hhole{\hhole{\hexp}}$. In our semantics given here, we eliminated them as they came up in a somewhat \emph{ad hoc} manner. We have not yet
explored an equational theory for terms with holes, but intend to once our
formalization effort is more mature.

@@

\bibliography{../research}
%\softraggedright
%P. Q. Smith, and X. Y. Jones. ...reference text...

>>>>

\def\OPTIONConf{0}%
%\documentclass[12pt]{article}
\documentclass[preprint,9pt]{sigplanconf}
% The following \documentclass options may be useful:
%
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{ stmaryrd }
\usepackage{mathpartir}
\usepackage{stmaryrd}
\usepackage{wasysym}
\usepackage{extarrows}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\usepackage{mathpazo}
\usepackage{colortab}
\usepackage{url}
\usepackage{todonotes}
\usepackage{listings}
\lstset{tabsize=2, 
basicstyle=\ttfamily\fontsize{8pt}{1em}\selectfont}
\usepackage{microtype}
\sloppy
\def \TirNameStyle #1{\small\rulename{#1}}
\renewcommand{\MathparLineskip}{\lineskiplimit=.3\baselineskip\lineskip=.3\baselineskip plus .2\baselineskip}

@@

\section{Hazelnut, Formally}
\label{sec:hazel}
\todo{revise text}\todo{add let and ifz}Hazelnut is based on the simply-typed lambda calculus extended with a single base type, $\tnum$. Its major constituents, introduced by example in the previous section, are:
\begin{itemize}
\item \textbf{H-types} and \textbf{H-expressions} (Sec. \ref{sec:holes}), which are terms with \emph{holes}. Holes mark subterms that are ``under construction.'' H-types classify H-expressions according to a {bidirectionally typed} static semantics.
\item \textbf{Z-types} and \textbf{Z-expressions} (Sec. \ref{sec:cursors}), which superimpose a single \emph{focus} onto H-types and H-expressions (using Huet's \emph{zipper pattern} \cite{JFP::Huet1997}.)
\item \textbf{Actions} (Sec. \ref{sec:actions}), which move the focus or modify the subterm in focus.

@@

We refer to terms that do not contain subterms of hole form as \emph{complete}. Informally, we will use metavariables $\tau$ and $e$ rather than $\htau$ and $\hexp$ for complete H-types and H-expressions, respectively. Formally, we can derive $\hcomplete{\tau}$ when $\tau$ is a complete H-type, and $\hcomplete{e}$ when $e$ is a complete H-expression. We omit the straightforward definitions of these judgements for concision. The dynamics of Hazelnut, which we need not detail here, is defined only  over complete H-expressions (i.e. we can only ``run'' a complete program, though see Sec. \ref{sec:future}.)\todo{get rid of e/edot distinction}\todo{change this to refer to dynamics section}

@@

The rules for relative movement within Z-expressions are similar. Movement is type-independent, so we defer to an auxiliary judgement for both the analytic and synthetic judgements:\todo{remove prevSib}
\begin{subequations}
\begin{equation}
\inferrule{
  \performMove{\zexp}{\aMove{\delta}}{\zexp'}
}{
  \performSyn{\hGamma}{\zexp}{\htau}{\aMove{\delta}}{\zexp'}{\htau}
}
\end{equation}
\begin{equation}
  \inferrule{
  \performMove{\zexp}{\aMove{\delta}}{\zexp'}
}{
  \performAna{\hGamma}{\zexp}{\htau}{\aMove{\delta}}{\zexp'}
}
\end{equation}
\end{subequations}
For concision, we show only the rules for ascription here:
\begin{subequations}
  \begin{equation}
    \label{r:movefirstchild}
  \inferrule{ }{
    \performTyp{
      \zwsel{\hexp : \htau}
    }{
      \aMove{\dChild}
    }{
      \zwsel{\hexp} : \htau
    }
  }
\end{equation}
\begin{equation}
  \label{r:moveparent}
  \inferrule{ }{
    \performTyp{
      \zwsel{\hexp} : \htau
    }{
      \aMove{\dParent}
    }{
      \zwsel{\hexp : \htau}
    }
  }
\end{equation}
\begin{equation}
  \inferrule{ }{
    \performTyp{
      \hexp : \zwsel{\htau}
    }{
      \aMove{\dParent}
    }{
      \zwsel{\hexp : \htau}
    }
  }
\end{equation}
\begin{equation}
  \label{r:movenextsib}
  \inferrule{ }{
    \performTyp{
      \zwsel{\hexp} : \htau
    }{
      \aMove{\dNext}
    }{
      \hexp : \zwsel{\htau}
    }
  }
\end{equation}
\begin{equation}
  \label{r:moveprevsib}
  \inferrule{ }{
    \performTyp{
      \hexp : \zwsel{\htau}
    }{
      \aMove{\dPrev}
    }{
      \zwsel{\hexp} : \htau
    }
  }
\end{equation}
\begin{equation}
\inferrule{
  \performTyp{
    \zexp
  }{
    \aMove{\delta}
  }{
    \zexp'
  }
}{
  \performTyp{
    \zexp : \htau
  }{
    \aMove{\delta}
  }{
    \zexp' : \htau
  }
}
\end{equation}
\begin{equation}
  \inferrule{
    \performTyp{
      \ztau
    }{
      \aMove{\delta}
    }{
      \ztau'
    }
  }{
    \performTyp{
      \hexp : \ztau
    }{
      \aMove{\delta}
    }{
      \hexp : \ztau'
    }
  }
\end{equation}
and this for the example\todo{revise}
\begin{equation}\label{r:movefirstchild-lam}
\inferrule{ }{
  \performMove{
    \zwsel{\hlam{x}{\hexp}}
  }{
    \aMove{\dChild}
  }{
    \hlam{x}{\zwsel{\hexp}}
  }
}
\end{equation}

@@

Reachability\todo{state reachability theorem}

@@

Constructibility\todo{state constructibility theorem}
\end{subequations}

@@

\section{Dynamics of Hazelnut}\label{sec:dynamics}

\subsection{Complete Expressions}
... canonical forms and type safety with e complete in it ...

\subsection{Holes as TODOs}
... add an e err and define new type safety ...

\subsection{Indeterminate Expressions}
\input{dynamics}

@@

\section{Extending Hazelnut}\label{sec:extending}
\todo{write this section} Add sum types.

H-expression

Z-expression + Erasure rules

Statics

Movement actions

Construct actions

@@

\section{Discussion \& Conclusion}
\label{sec:future}
\todo{revise}This paper presented Hazelnut, a type theoretic structure editor calculus. Our aim is to take a principled approach to its design by formally specifying its semantics, providing strong metatheoretic guarantees, mechanizing its semantics and metatheory in Agda and implementing it using the concepts of  functional reactive programming. As of this submission, we have achieved reasonable confidence in the formal system presented above, and have transitioned our focus toward the mechanization and implementation efforts. By the time of presentation, we anticipate having complete or nearly complete versions of these.

@@

It may also be possible to give a dynamics to incomplete expressions. Prior work on staged evaluation suggests that there may be a connection to modal logic, viewing holes as quantifying over all possible terms that may fill them \cite{DBLP:journals/jacm/DaviesP01}. In developing a dynamic semantics, we will also need to handle terms like $\hhole{\hehole}$ and
$\hhole{\hhole{\hexp}}$. In our semantics given here, we eliminated them as they came up in a somewhat \emph{ad hoc} manner. We have not yet
explored an equational theory for terms with holes, but intend to once our
formalization effort is more mature.\todo{fix bib}

@@

\bibliography{../research}}
%\softraggedright
%P. Q. Smith, and X. Y. Jones. ...reference text...

### 2016-06-22 implementation.tex
added TODOs

The key question that must be answered for any implementation strategy is: how do we model a stream of actions from a user? Let us assume that these actions are chosen (using some input device, preferably, a keyboard) from some ``palette'' that never presents the user with actions that are not semantically well-defined, according to the action semantics defined earlier.
%In a traditional editor, the input from the user is a stream of characters, and there are no guarantees that at any point that the program is syntactically well-formed, so the designer leaves editing as an .
%In contrast, in a structure editor, the input from the user is a stream of operations.
As such, each new action will ``atomically'' generate a new Z-expression. 
This insight leads us to conclude that a natural way to implement this editor would be using event-based Functional Reactive Programming~\cite{Wan:2000:FRP:349299.349331} (FRP).
Figure~\ref{fig:FRP} illustrates the concept of an FRP-based implementation of a  structure editor organized like Hazelnut.
The input from the user is a stream of actions.  Each action results in a change to the underlying abstract model (i.e., a new Z-expression is created after each action.)
Each model change results in an updated \emph{view} which is then presented to the user.  The user can then consider this new view when they choose a new action as input.

>>>>

\todo{make figure take up only one column}\todo{change terminology in figure}\todo{revise}
The key question that must be answered for any implementation strategy is: how do we model a stream of actions from a user? Let us assume that these actions are chosen (using some input device, preferably, a keyboard) from some ``palette'' that never presents the user with actions that are not semantically well-defined, according to the action semantics defined earlier.
%In a traditional editor, the input from the user is a stream of characters, and there are no guarantees that at any point that the program is syntactically well-formed, so the designer leaves editing as an .
%In contrast, in a structure editor, the input from the user is a stream of operations.
As such, each new action will ``atomically'' generate a new Z-expression. 
This insight leads us to conclude that a natural way to implement this editor would be using event-based Functional Reactive Programming~\cite{Wan:2000:FRP:349299.349331} (FRP).
Figure~\ref{fig:FRP} illustrates the concept of an FRP-based implementation of a  structure editor organized like Hazelnut.
The input from the user is a stream of actions.  Each action results in a change to the underlying abstract model (i.e., a new Z-expression is created after each action.)
Each model change results in an updated \emph{view} which is then presented to the user.  The user can then consider this new view when they choose a new action as input.

### 2016-06-22 intro.tex
added TODOs

%Although all programs can be represented as text, most text does not correspond to a syntactically well-formed and semantically well-defined program. 
Programmers and the tools that they use often encounter text that does not correspond to a meaningful program. This may be because the programmer is in the midst of a sequence of 
edit actions that leaves the text temporarily malformed or ill-typed, or because the programmer has made a mistake\todo{mention YoungSeok's data here?}. It is difficult for tools to help programmers determine and execute a corrective course of action in such situations (e.g. by providing \todo{cite study on benefits of syntax highlighting}syntax highlighting and semantics-aware code completion services\todo{cite something?}.) %because the language definition is silent %For example, a code editor cannot assist the programmer in these situations.

@@

\emph{Structure editors} have long promised to address this
problem by allowing programmers to edit the tree structure of a program, or some rich projection thereof, 
directly. Each edit action leaves the program being constructed in a
structurally well-formed state.\footnote{In addition to eliminating malformed edit states, it is also often promised that programs can be written more quickly using a structure editor. However, we will not discuss such ``edit costs'' here, because they depend on particular implementation details, e.g. whether a keyboard or a mouse is in use.} We give examples of notable structure editors in Sec. \ref{sec:rw}.

The problem is that it is still possible to leave the program in a semantically meaningless (i.e. undefined) state, because language definitions generally only give meaning to complete, well-typed terms. This makes it difficult for humans and tools to reason about types and binding during the development process.  %More sophisticated semantic reasoning principles thus remain .

@@

In this paper, we present our ongoing work on a minimal structure editor, Hazelnut,  defined in the type-theoretic style (i.e. Hazelnut is a \emph{structure editor calculus}.) In Hazelnut, expressions and types with \emph{holes} have a well-defined static semantics. Edit actions are type-aware and leave the program in both a structurally and semantically well-defined state. In fact, edit actions maintain an even stronger invariant -- when acting on an expression whose type is determined by its surroundings (e.g. an expression in function argument position), only edit actions consistent with that type are permitted (we will formally state this invariant in Sec. \ref{sec:actions}.) This does not imply that programs need to be constructed in a strictly outside-in manner, however, because an expression that has a type that is not yet consistent with its surroundings will automatically be put into a {hole} that defers consistency analysis until the hole is \emph{finished}.

@@

  \item We then formally define Hazelnut in Section \ref{sec:hazel} and state the important metatheoretic properties.

>>>>

%Although programs can be represented as text, most text does not correspond to a syntactically well-formed and semantically well-defined program. 
Most programming languages define a textual syntax. This allows programmers to use text editors and other standard text-based tools to construct and manipulate programs. While this is of substantial practical utility, it also introduces some fundamental complexity into the programming process.

Cognitively, it requires that programmers understand the subtleties of text parsing (e.g. the relative precedence of the various forms.) This can be particularly challenging for novice and end-user programmers.\todo{citation?}

It also complicates matters for tool designers because they might be confronted with text that does not correspond to a well-formed, meaningful program\todo{mention YoungSeok's data here}. This may be because the programmer is in the midst of a sequence of 
edit actions that leaves the text temporarily malformed or ill-typed, or because the programmer has made a mistake. The language definition is silent about these situations, so it is difficult for tools to help programmers determine and execute a corrective course of action (e.g. by providing \todo{cite study on benefits of syntax highlighting}syntax highlighting and semantics-aware code completion services\todo{cite something?}.)

@@

\emph{Structure editors} are program editors where every edit action leaves the program being manipulated in a 
structurally well-formed state.\footnote{In addition to eliminating malformed edit states, it is also often promised that programs can be written more quickly using a structure editor. However, we will not discuss such ``edit costs'' here, because they depend on particular implementation details, e.g. whether a keyboard or a mouse is in use.} Scratch etc. for novices. mbeddr for pros. Lamdu and Unison ongoing efforts with a functional foundation. We give other examples of notable structure editors in Sec. \ref{sec:rw}.\todo{revise}

The problem is that it is still possible to leave the program in a semantically meaningless (i.e. undefined) state, because language definitions generally only give meaning to complete, well-typed terms. This makes it difficult for humans and tools to reason about types and binding during the development process, even when using a structure editor. Some structure editors attempt stuff, but it is not clear what invariants are being maintained (e.g. Unison ; Scratch seems not to allow literals of the wrong type, but variables are not typed.)\todo{revise} %More sophisticated semantic reasoning principles thus remain .

@@

\todo{revise remainder of intro}In this paper, we present our ongoing work on a minimal structure editor, Hazelnut,  defined in the type-theoretic style (i.e. Hazelnut is a \emph{structure editor calculus}.) In Hazelnut, expressions and types with \emph{holes} have a well-defined static semantics. Edit actions are type-aware and leave the program in both a structurally and semantically well-defined state. In fact, edit actions maintain an even stronger invariant -- when acting on an expression whose type is determined by its surroundings (e.g. an expression in function argument position), only edit actions consistent with that type are permitted (we will formally state this invariant in Sec. \ref{sec:actions}.) This does not imply that programs need to be constructed in a strictly outside-in manner, however, because an expression that has a type that is not yet consistent with its surroundings will automatically be put into a {hole} that defers consistency analysis until the hole is \emph{finished}.

@@

  \item We then formally define Hazelnut in Section \ref{sec:hazel} and state the important metatheoretic properties.
  \item Dynamics \ref{sec:dynamics}
